\chapter*{Acknowledgment}
\chaptermark{Acknowledgment}
\addcontentsline{toc}{chapter}{\textit{Acknowledgment}} 

%\begin{spacing}{1.3}
All praise be to Allah. I thank Allah for providing me with the knowledge, capability, and perseverance to persist and complete this thesis. Among His means, firstly, I express my humble gratitude to my supervisor, Dr. M. Sohel Rahman, for his constant encouragement and guidance in completing this work. His patient hearing of my ideas, critical analysis of my observations, and detecting flaws (and amending thereby) in my thinking and writing have made this thesis successful. While pursuing my Ph.D. in Bangladesh, I faced unexpected social and financial challenges. Sometimes they seem more complicated to tackle than the technical challenges of this thesis. On several occasions, I felt to terminate my Ph.D. During those difficult days, I repeatedly learned that there is no power and no strength except with Allah. The fatherly support of my supervisor is a great blessing from Allah that helped me get out of my anxiety and mental pressures.

I want to thank the respected members of my thesis committee for their valuable suggestions. I thank my respected teachers Dr. Mahmuda Naznin, Dr. Md. Mostofa Akbar, Dr. M. Kaykobad, Dr. Md. Shamsuzzoha Bayzid, and Dr. Atif Hasan Rahman, from whom I have been learning since my undergraduate studies. I'm grateful to the external members Dr. Md. Tofazzal Islam and Dr. Swakkhar Shatabda. Their queries and comments have driven me to think about my thesis work differently.

I express my special gratitude towards Dr. Md. Shamsuzzoha Bayzid, Dr. Atif Hasan Rahman, and Dr. M. Saifur Rahman for the theoretical lessons and practical guidelines they extended to me regarding the field of computational biology throughout the whole period of this thesis. During my Ph.D. journey, two of my beloved colleagues, T.M. Tariq Adnan and Shehab Sarar Ahmed, acted like younger brothers. They provided me with crucial support whenever I needed them most. 

During this journey, my parents and wife always did their best to make things easier for me. My parents realized my financial challenges and acted proactively. My wife took care of a lot of household issues by herself so that I can focus on my research activities. I thank Allah to bless me with such an affectionate family. 
Finally, I would like to express my gratefulness to my elder brother, friends, and others who supported me during this journey. May Allah reward them all in here and hereafter.

%\end{spacing}